Sequence Specific Modeling of \emph{E.~coli} Cell-Free Protein Synthesis

### 2017-04-01 Paper_v1.tex
Working my way through the results

\section{Introduction}
Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamental biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
One tool to address this question is mathematical modeling.

@@

Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
Constraints based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions. 
These constraints based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} of arbitrary complexity, including genome scale networks, using linear programming\cite{Covert:2004aa}.  
Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
Stoichiometric reconstructions have been expanded to include metabolism integrated with detailed gene expression (ME-Model) \cite{Allen:2003aa, Brien693}, and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220} which has increased the scope of simulating genetic and environmental perturbations.
%However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}. 
Constraints based methods are great tools to estimate the performance of metabolic networks with none to very few parameters. 
However, constraints based methods are typically used to model \textit{in~vivo} processes and have not been applied to cell-free metabolism.

In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of promoter function, transcription and translation processes.
Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\subsection{Model derivation and validation}
The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
different cell-free \textit{E.~coli} extracts. The cell-free metabolic network, model parameters, model code,
and each protein sequence are available in the supplemental materials.

Cell-free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values.
On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
The cell-free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Fig-1-Network-Validation-Simulations.pdf}
\caption{Sequence specific flux balance analysis. A. Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway. Thick gray arrows indicate withdrawal of precursors for amino acid synthesis. B. CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption. Error bars denote the standard deviation of experimental measurements. C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption. Error bars denote a 10\% coefficient of variation. D. Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
\label{fig:network}
\end{figure}

Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.

\subsection{Analysis of CFPS performance}
Our next goal was to examine the performance of CFPS for eight different proteins under three different cases.
Each of the proteins was produced under a P70a promoter, except for CAT which was produced under a T7 promoter.
In all cases, CFPS was supplied with glucose.
In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (AA uptake and synthesis).
In the second case, CFPS was supplied with amino acids, but the amino acid synthesis reactions were turned off (AA uptake w/o synthesis).
These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).
Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
We used ssFBA to estimate the productivity, energy efficiency, and carbon yield for each of these proteins for each case.
An additional case was considered for CAT, since a comprehensive dataset is available (Supporting Information); in this case, fluxes were constrained to experimental measurements where avaliable, with the exception of CAT production which was determined by the transcription/translation parameters.

@@

\subsubsection{Productivity}
The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

@@

\subsubsection{Yield}
The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
The same inverse qualitative trend was observed as for energy efficiency.
%The cases supplemented with amino acids in the media showed the highest carbon yield.
There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.
%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.
Only the necessary amount of amino acids was used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
In that case, the carbon yield without glucose contribution would be 100\%.
%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.
This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
In addition, ssFBA is solved by maximizing the flux through the protein production reaction.
Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate, pyruvate and lactate.
The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
% \begin{figure}[t!]
% \centering
% \includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
% \caption{CFPS carbon yield of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency across the ensemble (error bars represent the 95\% CI). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for AA synthesis without uptake (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90). Dagger: constrained by experimental measurements and excluded from trendline.}
% \label{fig:Yield}
% \end{figure}
%Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
%The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}F).
The trendlines showed good predictability for estimating carbon yield for a certain range of potein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.
Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.

\subsection{Sensitivity analysis}
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for CAT, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield which followed the same trend of significance as energy efficiency (available in Supporting Information).
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}

>>>>

\section{Introduction}
Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamental biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
One tool to address these questions is mathematical modeling.

@@

Stoichiometric reconstructions of microbial metabolism, popularized by constraint based approaches such as flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
Constraint based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
Constraint based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} for biochemical networks of arbitrary complexity, including genome scale networks, using linear programming \cite{Covert:2004aa}.
Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B. subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
Stoichiometric reconstructions have been expanded to include the integration of with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693},
and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220}.
These expansion have greatly expanded the scope of questions these models can explore.
%However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}.
Constraint based methods are powerful tools to estimate the performance of metabolic networks with very few adjustable parameters.
However, constraint based methods are typically used to model \textit{in~vivo} processes, and have not yet been applied to cell-free metabolism.

In this study, we used sequence specific constraint based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of promoter function, transcription and translation processes.
Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.

@@

\subsection{Model derivation and validation}
The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa},
and adding deletions associated with the specific cell free system (see materials and methods).
We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
The network described the major carbon and energy pathways, as well as amino acid biosynthesis and degradation pathways.
Using this network, in combination with effective promoter models taken from Moon et al. [REF], and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
different \textit{E.~coli} cell-free extracts.
The cell free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [REF].

Cell free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig. ~\ref{fig:network}B and C).
Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress}
for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (Correlation?).
On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (Correlation?).
Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
transcription and translation elongation rates or the upper bounds for oxygen and glucose consumption rates (modeled as being normally distributed around the parameter values shown in Table ~\ref{tbl:parameters})
did not qualitatively alter the performance of the model (blue region, 95\% confidence estimate).
Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell free metabolic model, were consistent with experimental measurements.
However, these simulations were conducted at a single plasmid concentration.
Thus, it was unclear if the model could capture cell free protein synthesis for a range of plasmid concentrations.

\begin{figure}[t!]
\centering
\includegraphics[width=1.00\textwidth]{./figs/Fig-1-Network-Validation-Simulations.pdf}
\caption{Sequence specific flux balance analysis. A. Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway. Thick gray arrows indicate withdrawal of precursors for amino acid synthesis. B. CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption. Error bars denote the standard deviation of experimental measurements. C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption. Error bars denote a 10\% coefficient of variation. D. Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0. 95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
\label{fig:network}
\end{figure}

Cell free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
The cell free deGFP titer at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration.
However, while the mean and 95\% confidence estimate of the ensemble was consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.

%These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
% We used ssFBA to estimate the productivity, energy efficiency, and carbon yield for each of these proteins for each case.

\subsection{Analysis of CFPS performance}
To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell free production of eight model proteins with and without amino acid supplementation (Fig. \ref{fig:Prof}).
The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
In all cases, CFPS was supplied with glucose, however, we also considered different scenarios for amino acid supplementation.
First, the CFPS reaction was supplied with amino acids, and the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we would expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction.
Lastly, the system was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP),
prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).

@@

\subsubsection{Productivity}
The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
However, each individual protein had a different level of productivity.
For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
The relative high productivity of CAT was due to its T7 promoter.
CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.

@@

%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
%The cases supplemented with amino acids in the media showed the highest carbon yield.
%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.

\subsubsection{Yield}
The theoretical maximum mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
The same inverse qualitative trend was observed as for energy efficiency.
There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.
For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.
Only the necessary amount of amino acids was used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
In that case, the carbon yield without glucose contribution would be 100\%.
In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.
This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
In addition, ssFBA is solved by maximizing the flux through the protein production reaction.
Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate, pyruvate and lactate.
The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
% \begin{figure}[t!]
% \centering
% \includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
% \caption{CFPS carbon yield of eight proteins for four cases. Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white). A. Energy efficiency across the ensemble (error bars represent the 95\% CI). B. Mean energy efficiency versus the carbon number for each corresponding protein. Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for AA synthesis without uptake (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90). Dagger: constrained by experimental measurements and excluded from trendline.}
% \label{fig:Yield}
% \end{figure}
%Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
%The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Prof}F).
The trendlines showed good predictability for estimating carbon yield for a certain range of protein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.
% Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.

\subsection{Sensitivity analysis}
To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for CAT, a representative protein (Fig.~\ref{fig:SI}).
The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
In addition, Li et al. have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
\begin{figure}[t!]
\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray). Error bars represent a 95\% CI on the sensitivity index.}
\label{fig:SI}
\end{figure}
